### 2021-08-25 plos_latex_template.tex
rj + ast working on diversity, welcoming, accessibility, and spotlight rules

When thinking about diversity, each one of us can have different points of view about the meaning of the term, depending on where we come from. If you are reading this, you may be here because you already have some experience in Diversity, Inclusion, and Accessibility practices. Or you may want to improve diversity in your conference but do not know how or where to start. This article does not intend to be a comprehensive review of this topic --this would be impossible-- but will give some ideas pointing in key directions.

Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity,  geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few. While human diversity should be celebrated and respected in every way, the truth is that there are implicit hierarchies along these axes, and some statuses (e.g., cisgender, white, male, from the US or Europe) hold the privilege of being the default settings for which all systems --including conferences-- are consciously and unconsciously built. As a result, most people in the world fall into some category where they are part of a marginalized group --as opposed to a privileged one. The first step towards a diverse and inclusive conference is to recognize that people are diverse, that some groups face discrimination and might be underrepresented in your community and event \cite{timperleyHeMoanaPukepuke2020}.

@@

Think about the groups that are usually unsupported or discriminated against in your community or event, and about the meaning of "diverse" in your context. Imagine the result you would like to see if you succeeded in organizing a diverse conference according to your vision. Does this mean an even gender distribution in your speakers? Does this relate to people of all races -especially black people- being present as coordinators, presenters, and in the audience? Does this mean having LGBTQ+ friendly-spaces or community participation from key geographic regions? This vision should guide and help hold the organizing team accountable along the way.

To accomplish this diversity vision, you will have to take some conscious decisions toward removing bias from your conference's spaces. Having diverse people in decision-making positions will affect positively all the other aspects of your conference, not only internally because all the decision-making process will benefit from their input (see Rule \ref{rule_organizing_team}), but also externally because representation matters[see Rule \ref{rule_communication}].

@@

While you may want to improve representation towards some of the most visible dimensions of human diversity, such as race, gender, and country of origin, building a truly inclusive environment means taking care of all the other aspects as well. Having consideration of religious practices, setting specific accommodations for lactating women and child care, having LGBTIQ+-friendly spaces, creating community-only spaces, enforcing the use of pronouns, and treating gender as a non-binary variable, are just some examples of decisions that can make inclusion real.

@@

Some teams gather this vision in a diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}) to express the values of the conference and make diversity part of the outreach and communication strategy see Rule \ref{rule_communication}.

However, it is important to not stop with a statement alone, but to identify specific goals and tasks that ensure compliance with these larger goals, for example, identifying accessible platforms, developing a detailed code of conduct and a team to enforce it, and preparing accessibility guidelines (see other Rules below).

Lack of representation, unwelcoming -or overtly aggressive- environments hinder participation (or future participation) of people who could otherwise become active community members. In extreme cases, they can divert career paths, affect lives, and exclude people from some fields. So take this seriously, think about the reasons why you want to create a more diverse conference, and be open to learn. Do not treat diversity as a checklist, do not use people from minoritized groups for image purposes (tokenization), and do not act as a savior but as a genuine collaborator. %ast: some version of this might come in a box.

@@

Alternatively, a conference could have a hybrid format with an in-person and online component. This dual format could allow a group of people to interact face-to-face while providing many others the opportunity to participate remotely. The challenge in this kind of setting would be to make the online component as relevant as the in-person component and not just a consolation prize to the less privileged in the community \cite{ninerBetterWhomLeveling2021}.

\section{Rule X: Have an inclusive and diverse organizing team}
\label{rule_organizing_team}
A genuinely inclusive conference can only be organized by an inclusive and diverse organizing team. Build a team with people from different regions, genders, socioeconomic statuses, and other aspects of diversity. Particular attention should be paid to the usually marginalized groups (see \textbf{Rule 1} discussing the dimensions of diversity lacking in your community and event). To ensure a deeper understanding of the challenges in different diverse groups, it is essential to create a representative working group that functions as a snapshot of the community at large. But this will only work if there is real inclusion. One would be surprised by how often conference leads tend to do everything themselves instead of reaching out to the right people for the right task -- likely because the latter involves careful planning and communication well ahead of time or because they naively overestimate their expertise in areas they lack practical experience in. For example, a cisgender or abled person estimating what might work best for transgender people or people with sensory disabilities. The same holds for other dimensions of diversity. A truly inclusive and welcoming space is one in which everyone in the team is invited and allowed to bring their experience to bear. Even if creating and maintaining such a team and space is challenging, the positive outcomes far outweigh any minor inconveniences (e.g. time zones), and the team will grow, learn, and embrace more inclusive practices \cite{hongGroupsDiverseProblem2004}.

%The team may also have to review carefully and teach themselves the correct vocabulary for internal and external communication and identify the best ways to account for every culture and situation. % rj: left this out, Heather said it doesn't make much sense here %ast: we can collect part of this in rule 1, since some self-teaching, and not expecting to be taught, is necessary

%%Disabled people often repeat "Nothing about us without us", and this is true for every marginalized group. This means the actual life experiences, expertise, and insights from people in marginalized groups are not replaceable with good intentions from people outside these groups. 
%rj: Yani's refs to have just in case:
%https://sites.lsa.umich.edu/scottepage/wp-content/uploads/sites/344/2015/11/pnas.pdf; https://www.pnas.org/content/early/2014/11/13/1407301111; https://www.mckinsey.com/business-functions/organization/our-insights/why-diversity-matters

@@

Some practices, such as making accessible slides and presentations, are not yet common practice and will require great efforts from the presenters if they are not used to them. 
For that reason, the organizing team should provide accessibility guidelines for slides and presentations, encourage their use, and be available for any questions they may have.  
Among other things, the guidelines should ask for raw and accessible material to the talks before the conference, e.g., R Markdown, HTML, or \TeX{} files; if presentations are pre-recorded, the speakers should include their video and ensure that their face is visible so that deaf and hard-of-hearing people can read their lips if needed (see \url{https://user2021.r-project.org/participation/accessibility/} for example). 
Accessibility awards at the end of the conference are an excellent way to acknowledge the speakers who were mindful of inclusiveness when preparing and delivering their talks.

@@

\section{Rule X: Adopt a code of conduct and create a team to reinforce it}
\label{rule_CoC}

To ensure an inclusive, friendly, and safe space in the conference, the organizers need to adopt a code of conduct and set up a team to enforce it \cite{favaroYourScienceConference2016}. The code of conduct is a document meant to keep the community safe and should state clearly the unacceptable behaviors, the spaces of the conference in which it applies, the consequences for engaging in unacceptable behavior, and the way to report violations \cite{auroraHowRespondCode2018}. 
The code of conduct should be displayed prominently in several spaces of the conference to deter people from unacceptable behavior.

The code of conduct team should receive training on how to receive reports, respond to incidents, and communicate their responses. A diverse code of conduct team will be more understanding of intersectionality issues in discrimination and harassment practices. 
%The people who need the protection of a code of conduct are usually those with less power or privilege, as more powerful or privileged people are often already protected from most harm.'

@@

%rationale: 1. they might select for the already privileged or people who dont need those to be nice on the cvs 2. they select for outgoing extroverted privileged ("brave") people 3. they are time and emotionally consuming which selects out some people 4. they single out people. even if you controlled for 1 to 3 you would "badge" people - you are creating a minority while trying to avoid it.
Offering scholarships or grants to attend the conference may be counterproductive, as they can be seen as more of a competition and attract people that might not need the money (e.g., when their PI could pay for it) but could apply for it to enrich their CV while filtering out people who do not feel entitled to earn them. If the goal is to help the people who really need it, the language needs to make that clear without shrouding it with fancy-sounding phrases or terms that single out the recipients (e.g. `grant recipient' or `diversity scholar', respectively). Do not call it a scholarship or a grant. Ultimately, it is a fee waiver or a discount for the people who need additional financial support. Also, the process for applicants should be simplified. Applying for loans, grants, and scholarships may be an emotionally demanding task. Do not complicate the attendees' lives by asking for long essays to convince a committee that they deserve your support. For this stage of the process, a certain degree of trust goes a long way -- when they say they need the support, it is best to give them the waiver rather than second guess their eligibility. If the conference resources allow for it, you could even take further steps to offer financial support for activities that help them have the time and resources to be at the conference: child care support, transportation fees (if in-person), or internet connection services (if virtual).

@@

>>>>

Each one of us can have different points of view about the meaning of diversity, depending on where we come from. If you are reading this, you may be here because you already have some experience in Diversity, Inclusion, and Accessibility practices. Or you may want to improve diversity in your conference but do not know how or where to start. This article does not intend to be a comprehensive review of this topic --this would be impossible-- but will give some ideas pointing in key directions.    
% RJ: Most of the first paragraph could pivot to the intro, but we can see later.

Diversity encompasses multiple dimensions: age, physical ability, career stage, gender, gender identity,  geographic origin, language, neurodiversity, race, religion, sexual orientation, and socioeconomic background, to name a few. Human diversity should be celebrated and respected in every way. Nonetheless, there are implicit hierarchies along these axes, and some statuses (e.g., cisgender, white, male, from the US or Europe) hold the privilege of being the default settings for which all systems --including conferences-- are consciously and unconsciously built. As a result, most people in the world fall into some category where they are part of a marginalized group --as opposed to a privileged one. The first step towards a diverse and inclusive conference is to recognize that people are diverse, and that some groups face discrimination (some more than others) and might be underrepresented in your community and event \cite{timperleyHeMoanaPukepuke2020}.

@@

Think about the groups that are usually unsupported or discriminated against in your community or event, and about the meaning of "diverse" in your context. Imagine the result you would like to see if you succeeded in organizing a diverse conference according to your vision. Would this translate into an even gender distribution in your speakers? Would it be representation of all races -especially Black people-- among the head organizers, speakers, and attendees? Would it be having LGBTQ+ friendly-spaces or community participation from key geographic regions? This vision should guide and help hold the organizing team accountable along the way.

To accomplish this diversity vision, you will have to take conscious decisions toward removing bias from your conference's spaces. Having diverse people in decision-making positions will affect positively all the other aspects of your conference because all the decision-making processes will benefit from their input (see Rule \ref{rule_organizing_team}), and because representation matters (see Rule \ref{rule_communication}).
% RJ: we could improve the wording of the last sentence.

@@

While it is essential to improve representation towards some of the most visible dimensions of human diversity, such as race, gender, and country of origin, building a truly inclusive environment means taking care of all the other aspects of diversity as well. Having consideration of religious practices, setting specific accommodations for lactating women and child care, having LGBTIQ+-friendly spaces, creating community-only spaces, enforcing the use of pronouns, and treating gender as a non-binary variable are just some examples of decisions that can make inclusion real.

@@

%Some teams gather this vision in a diversity statement (e.g., \url{https://user2021.r-project.org/about/diversity_statement/}) to express the values of the conference and make diversity part of the outreach and communication strategy see Rule \ref{rule_communication}. However, it is important to not stop with a statement alone, but to identify specific goals and tasks that ensure compliance with these broader goals, for example, identifying accessible platforms, developing a detailed code of conduct and a team to enforce it, and preparing accessibility guidelines (see other Rules below).

Paying attention to some dimensions of diversity while neglecting others may have unintended consequences. Lack of representation, unwelcoming --or overtly aggressive-- environments hinder participation (or future participation) of people who could otherwise become active community members. In severe cases, they can divert career paths, affect lives, and exclude people from some fields.

%So take this seriously, think about the reasons why you want to create a more diverse conference, and be open to learn. Do not treat diversity as a checklist, do not use people from minoritized groups for image purposes (tokenization), and do not act as a savior but as a genuine collaborator. %ast: some version of this might come in a box.

A key component in creating a safe environment during a conference is adopting a code of conduct and creating and training a team to enforce it \cite{favaroYourScienceConference2016}. The code of conduct is a document meant to keep the community safe and should state clearly: the unacceptable behaviors, the spaces of the conference in which it applies, the consequences for engaging in unacceptable behavior, and the way to report violations \cite{auroraHowRespondCode2018}. 
The code of conduct should be displayed prominently in several spaces of the conference to deter people from unacceptable behavior. An efficient code of conduct acts as a protection for the community because the people who may be target of unacceptable behavior tend to be people with less power or privilege.

The code of conduct team should receive training on how to receive reports, respond to incidents, and communicate their responses. A diverse code of conduct team will be more understanding of power dynamics, and sensitive to discrimination and harassment. We strongly recommend reading How to enforce a Code of Conduct by Valerie Aurora (\url{https://frameshiftconsulting.com/resources/code-of-conduct-book/}) as an excellent starting point for the Code of Conduct team work.

@@

Alternatively, a conference could have a hybrid format with an in-person and online component. This dual format could allow a group of people to interact face-to-face while providing many others the opportunity to participate remotely. The challenge and requirement for this kind of setting would be to make the online component as relevant as the in-person component and not just a consolation prize to the less privileged in the community \cite{ninerBetterWhomLeveling2021}.

\section{Rule X: Have an inclusive and diverse organizing team}
\label{rule_organizing_team}
A genuinely inclusive conference can only be organized by an inclusive and diverse organizing team. Build a team with people from different regions, genders, socioeconomic statuses, and other aspects of diversity. Particular attention should be paid to the usually marginalized groups (see \textbf{Rule \ref{rule_diversity}} discussing the dimensions of diversity lacking in your community and event). To ensure a deeper understanding and smooth communication with different diverse groups, it is essential to create a representative working group that functions as a snapshot of the community at large. But this will only work if there is real inclusion. One would be surprised by how often conference leads tend to do everything themselves instead of reaching out to the right people for the right task -- likely because the latter involves careful planning and communication well ahead of time or because they naively overestimate their expertise in areas they lack practical experience in. For example, a cisgender or abled person estimating what might work best for transgender people or people with sensory disabilities. The same holds for other dimensions of diversity. A truly inclusive and welcoming space is one in which everyone in the team is invited and allowed to bring their experience to bear. Even if creating and maintaining such a team and space is challenging, the positive outcomes far outweigh any minor inconveniences (e.g. time zones), and the team will grow, learn, and embrace more inclusive practices \cite{hongGroupsDiverseProblem2004}.

%The team may also have to review carefully and teach themselves the correct vocabulary for internal and external communication and identify the best ways to account for every culture and situation. % rj: left this out, Heather said it doesn't make much sense here %ast: we can collect part of this in rule 1, since some self-teaching, and not expecting to be taught, is necessary. nope. now i think this is here.

%%Disabled people often repeat "Nothing about us without us", and this is true for every marginalized group. This means the actual life experiences, expertise, and insights from people in marginalized groups are not replaceable with good intentions from people outside these groups.

%rj: Yani's refs to have just in case:
%https://sites.lsa.umich.edu/scottepage/wp-content/uploads/sites/344/2015/11/pnas.pdf; https://www.pnas.org/content/early/2014/11/13/1407301111; https://www.mckinsey.com/business-functions/organization/our-insights/why-diversity-matters

@@

Some practices, such as making accessible slides and presentations, are not yet common practice and will require great efforts from the presenters if they are not used to them. 
For that reason, the organizing team should provide accessibility guidelines for slides and presentations, encourage their use, and be available for any questions they may have.  
Among other things, the guidelines should ask for raw and accessible material to the talks before the conference, e.g., R Markdown, HTML, or \TeX{} files; if presentations are pre-recorded, the speakers should include their video and ensure that their face is visible so that deaf and hard-of-hearing people can read their lips if needed (see \url{https://user2021.r-project.org/participation/accessibility/} for example).

@@

@@

Offering scholarships or grants to attend the conference may be counterproductive, as they can be seen as more of a competition and attract people that might not need the money (e.g., when their PI could pay for it) but could apply for it to enrich their CV while filtering out people who do not feel entitled to earn them. If the goal is to help the people who really need it, the language needs to make that clear without shrouding it with fancy-sounding phrases or terms that single out the recipients (e.g. `grant recipient' or `diversity scholar', respectively). Do not call it a scholarship or a grant. Ultimately, it is a fee waiver or a discount for the people who need additional financial support. Also, the process for applicants should be simplified. Applying for loans, grants, and scholarships may be an emotionally demanding task. Do not complicate the attendees' lives by asking for long essays to convince a committee that they deserve your support. For this stage of the process, a certain degree of trust goes a long way -- when they say they need the support, it is best to give them the waiver rather than second guess their eligibility. If the conference resources allow for it, you could even take further steps to offer financial support for activities that help them have the time and resources to be at the conference: child care support, transportation fees (if in-person), or internet connection services (if virtual).

@@

% RJ: could add some ideas from here 
% https://www.science.org.au/files/userfiles/support/emcr/documents/one-page-summary-emcr-improving-diversity-web.pdf
